---
title: "NIH - Mali TBV Bayesian Models"
output: html_notebook
---

Emily Reichert, July 2019
This Rmd file takes the cleaned study datasets from Mali, Uganda, and Myanmar and generates Bayesian logistic regression models for probability of uRDT detection based on HRP2 (pg/ml) concentrations. (i.e. generates Figure 2)

Part I. Fit Bayesian Logistic Probability of Detection Curves to Mali, Uganda, Myanmar
Note: First run 'mali_tbv_uganda_myanmar.R' and 'mali_tbv_clinical.Rmd' scripts to clean datasets 

```{r}
#load necessary libraries
library(dplyr)
library(binom)
library(foreign)
library(readstata13)
library(RColorBrewer)
library(brms)
```

A. Fit brms model to BIOME_1_ Uganda dataset (cleaned)
```{r}
#pull out HRP2 and uRDT variables
z3 <- BIOME1_uganda_1_ %>%
  select(HRP2_pg_ml, uRDT_result = HS_RDT_Result)

#fit model
library(rstan)
mod = brms::brm(
  uRDT_result ~ log10(HRP2_pg_ml),
  data = z3,
  family = "bernoulli",
  prior = prior('normal(0,3)'),
  iter = 1000,
  chains = 4,
  cores = 4
)
```

```{r}
hrp2.seq = seq(-1,5, by = 0.1)

newdata = expand.grid(HRP2_pg_ml = 10^hrp2.seq)

fit = fitted(
  mod,
  newdata = newdata,
  summary = TRUE
) * 100

colnames(fit) = c('fit','se','lwr','upr')
df1 = cbind(newdata,fit)
```

B. Fit brms model to SMRU Myanmar data (Cleaned)
```{r}
z5 = SMRU_Myanmar %>%
  select(HSRDT_num, HRP2_pg_ml)

mod = brms::brm(
  HSRDT_num ~ log10(HRP2_pg_ml),
  data = z5,
  family = "bernoulli",
  prior = prior('normal(0,3)'),
  iter = 1000,
  chains = 4,
  cores = 4
)
```

```{r}
hrp2.seq = seq(-1,5, by = 0.1)

newdata = expand.grid(HRP2_pg_ml = 10^hrp2.seq)

fit = fitted(
  mod,
  newdata = newdata,
  summary = TRUE
) * 100

colnames(fit) = c('fit','se','lwr','upr')
df3 = cbind(newdata,fit)
```

C. Fit brms model to Mali uRDT, NIH dataset (cleaned)
```{r}
z4 <- NIH %>%
  select(uRDT_result, HRP2_pg_ml)

mod = brms::brm(
  uRDT_result ~ log10(HRP2_pg_ml),
  data = z4,
  family = "bernoulli",
  prior = prior('normal(0,3)'),
  iter = 1000,
  chains = 4,
  cores = 4
)
```

```{r}
hrp2.seq = seq(-1,5, by = 0.1)
newdata = expand.grid(HRP2_pg_ml = 10^hrp2.seq)

fit = fitted(
  mod,
  newdata = newdata,
  summary = TRUE
) * 100

colnames(fit) = c('fit','se','lwr','upr')
df = cbind(newdata,fit)
```

D. Fit brms model to Mali co-RDT
```{r}
z1 = subset(NIH, !is.na(RDT))
z1 <- z1 %>%
  select(RDT, HRP2_pg_ml)

mod = brms::brm(
  RDT ~ log10(HRP2_pg_ml),
  data = z1,
  family = "bernoulli",
  prior = prior('normal(0,3)'),
  iter = 1000,
  chains = 4,
  cores = 4
)
```

```{r}
hrp2.seq = seq(-1,5, by = 0.1)

newdata = expand.grid(HRP2_pg_ml = 10^hrp2.seq)

fit = fitted(
  mod,
  newdata = newdata,
  summary = TRUE
) * 100

colnames(fit) = c('fit','se','lwr','upr')
df7 = cbind(newdata,fit)
```


```{r}
#make composite figure showing probability of detection curves for all study cohorts
pdf(file = "Mali_brms.pdf", width = 6.8, height = 5.9)

p <- plot(log10(df$HRP2_pg_ml), df$fit, col = "deeppink", lwd=2,type="l", xaxt = "n", yaxt = "n", ylab = "Probability of Detection by RDT (%)", xlab = expression(paste("HRP2 pg/ml")), xlim = c(-1,5), ylim = c(0,100))

#Add curve for Mali uRDT
abline(v = log10(c(1,10,10^2,10^3,10^4)), col = "darkgrey", lty = 5)
abline(h = seq(0,100, by=20), col="grey", lty=3)
polygon(c(log10(df$HRP2_pg_ml), rev(log10(df$HRP2_pg_ml))), c(df$lwr, rev(df$upr)), col=adjustcolor("deeppink", 0.3), border = NA)

#Add curve for Uganda uRDT
lines(log10(df1$HRP2_pg_ml), df1$fit, col = "deepskyblue", lwd=2,type="l", xaxt = "n", yaxt = "n", xlim = c(-1,6), ylim = c(0,100))
polygon(c(log10(df1$HRP2_pg_ml), rev(log10(df1$HRP2_pg_ml))), c(df1$lwr, rev(df1$upr)), col=adjustcolor("deepskyblue", 0.3), border = NA)

#Add curve for Myanmar uRDT
lines(log10(df3$HRP2_pg_ml), df3$fit, col = "darkorchid1", lwd=2,type="l", xaxt = "n", yaxt = "n", xlim = c(-1,6), ylim = c(0,100))
polygon(c(log10(df3$HRP2_pg_ml), rev(log10(df3$HRP2_pg_ml))), c(df3$lwr, rev(df3$upr)), col=adjustcolor("darkorchid1", 0.3), border = NA)

#Add curve for Mali coRDT
lines(log10(df7$HRP2_pg_ml), df7$fit, col = "orange", lwd=2,type="l", xaxt = "n", yaxt = "n", xlim = c(-1,6), ylim = c(0,100))
polygon(c(log10(df7$HRP2_pg_ml), rev(log10(df7$HRP2_pg_ml))), c(df7$lwr, rev(df7$upr)), col=adjustcolor("orange", 0.3), border = NA)

legend(-1,100,legend=c("Mali uRDT","Uganda uRDT","Myanmar uRDT", "Mali co-RDT"), col=c("deeppink","deepskyblue","darkorchid1", "orange"), lty = c(1,1,1), ncol=1)

axis(1,log10(c(10^0, 10^1, 10^2, 10^3, 10^4)), c("1", "10", "100", "1,000", "10,000"))
axis(2, seq(0,100, by=20), las = 1)

dev.off()
```


PART II. HRP2 distributions density plot

Arrange data into format for density plot
```{r}
a <- NIH[,14]
a$Population <- "Mali (n = 622, high transmission)"
b <- BIOME1_uganda_1_[,11]
b$Population <- "Uganda (n = 607, high transmission)"
c <- smru_myanmar[,26]
c$Population <- "Myanmar (n = 1845, low transmission)"

compiled_hrp2 = rbind(a,b,c)
```

Create plot comparing HRP2 distributions of three study populations
```{r}
compiled_hrp2$HRP2_pg_ml <- as.numeric(compiled_hrp2$HRP2_pg_ml)

ggplot(compiled_hrp2,aes(x=log10(HRP2_pg_ml),fill=Population, color = Population)) +
  geom_density(aes(y = ..scaled..), alpha = 0.3)+
  facet_wrap(~Population,nrow=3) +
  labs(y="Scaled density")+
  labs(x="HRP2 (pg/ml)") + scale_x_continuous(breaks = c(0,1,2,3,4,5), labels = c(10^seq(0,5))) + scale_fill_manual(values = c("deeppink", "darkorchid1", "deepskyblue")) + scale_color_manual(values = c("deeppink", "darkorchid1", "deepskyblue")) + theme(legend.position = "none", text = element_text(size = 14))
```

